---
title: "Alerts"
output: pdf_document
date: "2024-11-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load necessary libraries
```{r}
library(dplyr)
library(ggplot2)

titanic_data <- read.csv("~/Desktop/School/Data Analysis/Assignments/train.csv")  
```

#Overall survival Rate
```{r}
total_survived <- sum(titanic_data$Survived)
total_population <- nrow(titanic_data)
overall_survival_rate <- total_survived / total_population
overall_survival_rate
```
#.3838384 Overall Sruvival Rate

#female survival rate
```{r}
female_data <- filter(titanic_data, Sex == "female")
female_survived <- sum(female_data$Survived)
total_females <- nrow(female_data)
female_survival_rate <- female_survived / total_females
female_survival_rate
```
#.7420382 Female Survival rate

# Hypothesis: Is the female survival rate is higher than the overall survival rate
```{r}
result <- prop.test(female_survived, total_females, p = overall_survival_rate, conf.level = 0.95)
result
```
The test statistic is 168.84, the value highlights a substantial deviation of the female survival rate from overall rate.
THe P-Value is < 2.2e-16, which is far below the .05 significance level.
Our confidence level is between .6978206 and .78877
Our results suggest that the female survival rate is significantly higher than the overall rate.
We reject the null hypothesis, as there is strong evidence pointing towards the idea that the female survival rate is much higher than the overall rate.

#Graph
```{r}
Graph<- ggplot() +
  geom_point(aes(x = 1, y = female_survival_rate), color = "blue", size = 3) +
  geom_errorbar(aes(x = 1, ymin = result$conf.int[1], ymax = result$conf.int[2]), width = 0.1, color = "blue") +
  geom_hline(aes(yintercept = overall_survival_rate), color = "red", linetype = "dashed") +
  xlim(0.5, 1.5) +
  ylim(0, 1) +
  
  labs(title = "Comparison of Female and Overall Survival Rates", y = "Survival Rate", x = "") 
Graph
```


```{r Graph, echo=FALSE}
plot(Graph)
```
