---
title: "Fall 2024 Project - Group 2 Working Document"
author: "Brian, Angelo, Ryan, & Christina"
date: "Fall 2024"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE) # set this so that warnings do not print along with our output
```

```{r, warning = FALSE}
# first load libraries
library(readr)
library(dplyr)
library(psych)
library(ggplot2)
library(lubridate)
```

Our goals for checkpoint 1 include (1) select the data sets from the MBTA's official data portal, (2) Clean and merge these datasets, and (3) conduct some exploratory data analysis on each of the data sets separately, as well as on the merged set.

## Introduction to our Datasets

Our analysis uses publically available data from the MBTA's Blue Book Open Data Portal (https://mbta-massdot.opendata.arcgis.com). First, we will look at the ridership data - specifically the MBTA historical gate station entries (https://mbta-massdot.opendata.arcgis.com/datasets/7859894afb5641ce91a2bb03599fdf5b/about). Then, we will analyze this against the service alerts (https://mbta-massdot.opendata.arcgis.com/datasets/90ed9092bd7a4285b296d5ff938edf29_0/explore). 

The goal of this analysis will be to find the most impactful service alerts, whether that be type of service issue, particular line of the issue, or circumstances around the issues (i.e. were these unexpected or planned service delays/outages). From there we can provide recommendations to the MBTA on where to focus in the coming year.

## Dataset 1: Ridership Data

For the purpose of our initial analysis, we have looked at 3 years of recent ridership data. This data includes the number of gated entries at each MBTA T station by every half hour. Variables in this dataset include the date and timestamp, a stop_id identifying the unique MBTA T stop, station_name, route_or_line telling us which T line this was on, and the count of gated entries in that given half hour time period.

```{r}

# upload each file by year - change paths as applicable
file_2023 <- "/Users/christinawright/Documents/Fall 2024 Datasets/GSE_by_year/GSE_2023.csv"
file_2022 <- "/Users/christinawright/Documents/Fall 2024 Datasets/GSE_by_year/GSE_2022.csv"  # Update this path to where you have saved the file
file_2021 <- "/Users/christinawright/Documents/Fall 2024 Datasets/GSE_by_year/GSE_2021.csv"  # Update this path to where you have saved the file

# import the data
data_2023 <- read_csv(file_2023, show_col_types = FALSE)
data_2022 <- read_csv(file_2022, show_col_types = FALSE)
data_2021 <- read_csv(file_2021, show_col_types = FALSE)

# add a column to each dataset for the corresponding year
data_2023 <- data_2023 %>% mutate(Year = 2023)
data_2022 <- data_2022 %>% mutate(Year = 2022)
data_2021 <- data_2024 %>% mutate(Year = 2021)

# combine
Ridership_3yeardata <- bind_rows(data_2021, data_2022, data_2023)
```

```{r}
# preview the first 10 rows of the dataset
head(Ridership_3yeardata, 10)
```

```{r, warning=FALSE}
# Summary statistics 
summary_stats <- describe(Ridership_3yeardata)
# only stats I want to print
summary_stats <- summary_stats[, c("n", "mean", "sd" ,"max", "min")]
print(summary_stats)
```

```{r}
Ridership_3yeardata <- Ridership_3yeardata %>%
  mutate(service_date = as.Date(service_date),
         gated_entries = ifelse(gated_entries < 0, NA, gated_entries))  # Replace negative entries with NA
```

```{r}
Ridership_3yeardata <- Ridership_3yeardata %>%
  mutate(service_date = as.Date(service_date),  # Convert notif_start to Date
         month = floor_date(service_date, "month"),
         year = floor_date(service_date, "year") )  # Extract month for aggregation
```



### Dataset 2: Service Alert Data

Within our Service Alerts dataset, we have alert level data from the last 5 years. For the purpose of our analysis, we will likely filter this down to the last ~3 years, to match the time-span of our ridership data. Variables within this data set include the type of transportation (gui_mode_name), the type of alert (alert_type_name), the type of alert (effect_name), the cause of the effect (cause_name), the severity of the alert (severity_name & severity_code), some description / context behind the alert (header & description), and then corresponding timestamps for when the alerts start or end.

```{r}
# change path when downloaded to the location of the dataset
service_alerts <- read_csv('/Users/christinawright/Documents/Fall 2024 Datasets/MBTA_Service_Alerts.csv', show_col_types = FALSE)

# preview the first 10 rows of the dataset
head(service_alerts, 10)
```

```{r}
service_alerts <- service_alerts %>%
  mutate(notif_start = as.Date(notif_start),  # Convert notif_start to Date
         month = floor_date(notif_start, "month"),
         year = floor_date(notif_start, "year") )  # Extract month for aggregation
```


```{r}
summary_stats <- describe(service_alerts)
# only stats I want to print
summary_stats <- summary_stats[, c("n", "mean", "sd" ,"max", "min")]
print(summary_stats)
```

### Merging the two datasets: service alerts by subway line, along with ridership data

To ultimately be able to conduct an analysis on the relationship between service alerts and ridership, we will need the datasets to be merged into one clean and easy to use dataframe. To do so, we filter the sets to the same time periods, and join them on the T line and month-year level.

Create a column with only the color of the line in the ridership data
```{r}
# Extract the color before "line"
Ridership_3yeardata$SubwayLine <- toupper(sub(" Line", "", Ridership_3yeardata$route_or_line))
```

Filter to subway alerts only and create 4 new columns with flags for whether or not the alert message pertains to each line
```{r}
filtered_service_alerts <- service_alerts %>%
  filter(gui_mode_name == "Subway" & year(service_alerts$notif_start) %in% c(2021,2022, 2023))

# Add a column for each color with boolean 1 or 0 based on whether the description contains the color and "experiencing"
filtered_service_alerts$IsGreenLine <- ifelse(filtered_service_alerts$color == '00843D', 1, 0)
filtered_service_alerts$IsRedLine <- ifelse(filtered_service_alerts$color == 'DA291C', 1, 0)
filtered_service_alerts$IsBlueLine <- ifelse(filtered_service_alerts$color == '003DA5', 1, 0)
filtered_service_alerts$IsOrangeLine <- ifelse(filtered_service_alerts$color == 'ED8B00', 1, 0)
filtered_service_alerts$year = year(filtered_service_alerts$notif_start)
filtered_service_alerts$month = month(filtered_service_alerts$notif_start)
```

Aggregate the alert frequencies and serverity of alert by T line, for each month / year
```{r}
aggregated_service_alerts <- filtered_service_alerts %>%
  mutate(SubwayLine = case_when(
    IsGreenLine == 1 ~ "GREEN",
    IsRedLine == 1 ~ "RED",
    IsBlueLine == 1 ~ "BLUE",
    IsOrangeLine == 1 ~ "ORANGE",
    TRUE ~ NA_character_ # If none of the conditions match
  )) %>%
  group_by(year, month, SubwayLine) %>%
  summarise(
    distinct_alert_count = n_distinct(alert_id),  # Count distinct alert_ids
    avg_severity = mean(severity_code, na.rm = TRUE),  # Calculate average severity
    .groups = "drop"
  )
```

Aggregate the ridership data on the same level
```{r}
# Aggregate by year, month, and color, summing the gated_entries
aggregated_Ridership_data <- Ridership_3yeardata %>%
  group_by(Year, month, SubwayLine) %>%
  summarise(
    total_gated_entries = sum(gated_entries, na.rm = TRUE),
    .groups = "drop"
  )

aggregated_Ridership_data$month = month(aggregated_Ridership_data$month)
aggregated_Ridership_data$year = (aggregated_Ridership_data$Year)
```

Join the two aggregated data sets
```{r}
Alerts_Ridership_BySubwayLine <- inner_join(aggregated_Ridership_data, aggregated_service_alerts, 
                         by = c("year", "month", "SubwayLine"))

# preview the first 10 rows of the dataset
# Drop the duplicate 'Year' column
Alerts_Ridership_BySubwayLine <- Alerts_Ridership_BySubwayLine %>%
  select(-Year)
head(Alerts_Ridership_BySubwayLine, 10)
```

### Exploratory Data Analysis for Ridership Data

Looking at ridership data by MBTA T line, we can see the red line has lead in gated entries, followed by the orange, green, blue, then silver. The mattapan line is likely a misclassification of the red or orange lines, but we will investigate that further.
```{r}
# aggregate the data bythe T line variable, route_or_line
T_line_ridership <- Ridership_3yeardata %>%
  group_by(route_or_line) %>%
  summarize(total_gateentries = sum(gated_entries, na.rm = TRUE))

# plot the sum of entries by T line
ggplot(T_line_ridership, aes(x = reorder(route_or_line, -total_gateentries), y = total_gateentries, fill = route_or_line)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Total Ridership by MBTA T Line",
    x = "T Line",
    y = "Total Gated Entries"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

We can also see how ridership may have changed over time
```{r}
# adding in a 'month' column based off the service_date
Ridership_3yeardata <- Ridership_3yeardata %>%
  mutate(month = floor_date(service_date, "month")) 

# agreagte the data by month
monthly_ridership <- Ridership_3yeardata %>%
  group_by(month) %>%
  summarize(total_gateentries = sum(gated_entries, na.rm = TRUE))

# plotting the total ridership over time by month
ggplot(monthly_ridership, aes(x = month, y = total_gateentries)) +
  geom_line(color = "blue", size = 1) +
  scale_x_date(
    date_labels = "%b %Y",  # month year x-axis labels
    date_breaks = "1 month" # adding month grid lines for readability
  ) +
  labs(
    title = "Monthly Total Ridership Over Time",
    x = "Month",
    y = "Total Gated Entries"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

```

### Exploratory Data Analysis for Service Alerts Data

Within this section, we have plotted frequency of alerts of alert type, along with the severity of these alerts, and the duration of them.


```{r} 
# Find the 5 most frequent effect types
top_effects <- service_alerts %>%
  count(effect_name) %>%
  top_n(5, wt = n) %>%
  pull(effect_name)

# Calculate the average severity code for all top 5 effects
top_effects_data <- subset(service_alerts, effect_name %in% top_effects)

# Group and calculate the average severity code
average_severity <- top_effects_data %>%
  group_by(effect_name) %>%
  summarise(average_severity = mean(severity_code, na.rm = TRUE))

print(average_severity)
```

Visualizing this data
```{r}
# Data for visualization
effect_data <- data.frame(
  Effect_Name = c("Detour", "Delay", "Escalator Closure", "Elevator Closure", "Track Change"),
  Severity_Code = c(6.33, 4.97, 2.99, 2.97, 1.65)
)

# Create the bar chart
ggplot(effect_data, aes(x = reorder(Effect_Name, -Severity_Code), y = Severity_Code, fill = Effect_Name)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 5 Effect Names and Their Severity Code",
       x = "Effect Name",
       y = "Average Severity Code") +
  theme_minimal()
```

This next chart shows that Delay alerts are by far the most frequent, 
making them a key area for the MBTA to address. The high frequency of delays 
suggests that focusing on reducing these disruptions could significantly 
improve rider experience. Other alerts like Detour, Elevator Closure,
and Track Change, though less frequent, also impact riders, particularly 
in accessibility and route consistency. To optimize service, MBTA could analyze
these alerts by line or station, differentiate between planned and unplanned 
disruptions, and prioritize maintenance to prevent unexpected outages. For the
coming year, targeted efforts on reducing delays, enhancing communication, and
ensuring accessibility would likely yield the most positive impacts for riders.

```{r}
service_alerts <- service_alerts %>%
  mutate(notif_start = as.Date(notif_start),  # Convert notif_start to Date
         month = floor_date(notif_start, "month"))  # Extract month for aggregation

# Top Alert Frequency
top_alert_types <- service_alerts %>%
  count(effect_name) %>%
  arrange(desc(n)) %>%
  slice(1:5) %>%
  pull(effect_name)

# Only Top Alerts
filtered_alerts <- service_alerts %>%
  filter(effect_name %in% top_alert_types)

# Top Alert Count
monthly_alerts <- filtered_alerts %>%
  group_by(month, effect_name) %>%
  summarise(alert_count = n(), .groups = "drop")

# PAST 3 Years
start_date <- as.Date("2021-01-01")
monthly_alerts <- monthly_alerts %>% filter(month >= start_date)

#VIZ
ggplot(monthly_alerts, aes(x = month, y = alert_count, fill = effect_name)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Monthly Frequency of All Top Alert Types (Last 3 Years)",
       x = "Month",
       y = "Alert Frequency",
       fill = "Alert Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r}
severity_type <- unique(service_alerts$alert_time_type)

severity_counts <- service_alerts %>%
  count(alert_time_type, name = "frequency")


ggplot(severity_counts, aes(x = severity_type, y = frequency,  fill= severity_type))+
  geom_bar(stat = "Identity", fill = "skyblue", color = "white") +
  labs( title = "Count of Alert Time Type 2023", x = "Alter Time Type", y = "Count")+
  scale_y_continuous(labels = scales::label_number()) +
  theme_minimal()

```

### Exploratory Data Analysis for the Merged Dataset with both Ridership & Service Alerts

