---
title: "Fall 2024 Project - Group 2 Working Document"
author: "Brian, Angelo, Ryan, & Christina"
date: "Fall 2024"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE) # set this so that warnings do not print along with our output
```

```{r, warning = FALSE}
# first load libraries
library(readr)
library(dplyr)
library(psych)
library(ggplot2)
library(lubridate)
library(scales)
```

### Introduction to our Datasets

Our analysis uses publicly available data from the MBTA's Blue Book Open Data Portal (https://mbta-massdot.opendata.arcgis.com). First, we will look at the ridership data - specifically the MBTA historical gate station entries (https://mbta-massdot.opendata.arcgis.com/datasets/7859894afb5641ce91a2bb03599fdf5b/about). Then, we will analyze this against the service alerts (https://mbta-massdot.opendata.arcgis.com/datasets/90ed9092bd7a4285b296d5ff938edf29_0/explore). 

The goal of this analysis is to find the most impactful service alerts, whether that be type of service issue, particular line of the issue, or circumstances around the issues (i.e. were these unexpected or planned service delays/outages). From there we can provide recommendations to the MBTA on where to focus in the coming year.

### Dataset 1: Ridership Data

For the purpose of our initial analysis, we have looked at 3 years of recent ridership data. This data includes the number of gated entries at each MBTA T station by every half hour. Variables in this dataset include the date and timestamp, a stop_id identifying the unique MBTA T stop, station_name, route_or_line telling us which T line this was on, and the count of gated entries in that given half hour time period.

```{r}
RidershipDataPath <- "//Users/christinawright/Documents/GitHub/Fall-2024-Group-2/Datasets/Monthly_Ridership.csv"  # Update this path to where you have saved the file

# import the data
RidershipData <- read_csv(RidershipDataPath, show_col_types = FALSE)
```

We can preview the first 10 lines and see the following variables: Year, Month, SubwayLine, and Total_Gated_Entries

```{r}
head(RidershipData, 10)
```

```{r, warning=FALSE}
# Summary statistics 
summary_stats <- describe(RidershipData$total_gated_entries)
# only stats I want to print
summary_stats <- summary_stats[, c("n", "mean", "sd" ,"max", "min")]
print(summary_stats)
```

### Dataset 2: Service Alert Data

Within our Service Alerts dataset, we have alert level data from the last 5 years. For the purpose of our analysis, we filtered this down to the last ~3 years, to match the time-span of our ridership data. Variables within this data set include the type of transportation (gui_mode_name), the type of alert (alert_type_name), the type of alert (effect_name), the cause of the effect (cause_name), the severity of the alert (severity_name & severity_code), some description / context behind the alert (header & description), and then corresponding timestamps for when the alerts start or end.

First, we can look at the alert level table, which we use for deep diving into the root causes of alerts.
```{r}
ServiceAlertPath <- "/Users/christinawright/Documents/GitHub/Fall-2024-Group-2/Datasets/ServiceAlerts_AlertLevel_BySubwayLine.csv"  # Update this path to where you have saved the file

# import the data
ServiceAlertLevelData <- read_csv(ServiceAlertPath, show_col_types = FALSE)
```

We can see within this table, we have many columns relevant to the details of the alerts, with flags for which subway line it is impacting.

```{r}
head(ServiceAlertLevelData)
```

### Dataset 3: Monthly Service Alerts

Below is our monthly service alerts data

```{r}
MonthlyServiceAlertPath <- "/Users/christinawright/Documents/GitHub/Fall-2024-Group-2/Datasets/Monthly_ServiceAlerts.csv"  # Update this path to where you have saved the file

# import the data
MonthlyServiceAlertLevelData <- read_csv(MonthlyServiceAlertPath, show_col_types = FALSE)
```

We can see below that for every month & subway line, there is a distinct count of the number of service alerts, as well as the averager severity for those alerts during the given month.
```{r}
head(MonthlyServiceAlertLevelData)
```

### Dataset 4: Monthly Ridership & Service Alerts - Merged Dataset

To ultimately be able to conduct an analysis on the relationship between service alerts and ridership, we will need the datasets to be merged into one clean and easy to use dataframe. To do so, we filter the sets to the same time periods, and join them on the T line and month-year level.

```{r}
MonthlyAlertsRidershipPath <- "/Users/christinawright/Documents/GitHub/Fall-2024-Group-2/Datasets/Alerts_Ridership_BySubwayLine.csv"  # Update this path to where you have saved the file

# import the data
MonthlyAlertsRidershipData <- read_csv(MonthlyAlertsRidershipPath, show_col_types = FALSE)
```

Below, we can see that this table contains the monthly gated entries, alert count, and average severity for each of the subway lines.
```{r}
head(MonthlyAlertsRidershipData)
```

### Exploratory Data Analysis for Ridership Data

Looking at ridership data by MBTA T line, we can see the red line has lead in gated entries, followed by the orange, green, blue, then silver. The mattapan line is likely a misclassification of the red or orange lines, but we will investigate that further.
```{r}
# aggregate the data bythe T line variable, route_or_line
T_line_ridership <- RidershipData %>%
  filter(SubwayLine != "MATTAPAN") %>% # Exclude MATTAPAN
  group_by(SubwayLine) %>%
  summarize(total_gateentries = sum(total_gated_entries, na.rm = TRUE))

# plot the sum of entries by T line

ggplot(T_line_ridership, aes(x = reorder(SubwayLine, -total_gateentries), y = total_gateentries, fill = SubwayLine)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = scales::label_number(scale = 1e-6, accuracy = 0.1)(total_gateentries)), 
            vjust = -0.5, color = "black") + # format labels in millions
  scale_fill_manual(
    values = c(
      "RED" = "red",
      "BLUE" = "blue",
      "GREEN" = "green",
      "ORANGE" = "orange",
      "SILVER" = "darkgray"
    )
  ) + 
  labs(
    title = "Total Ridership by MBTA T Line (millions)",
    subtitle = "Ridership from 2021-2024",
    x = "T Line",
    y = "Total Gated Entries (millions)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")


```


### Exploratory Data Analysis for Service Alerts Data

Within this section, we have plotted frequency of alerts of alert type, along with the severity of these alerts, and the duration of them.

```{r}
# aggregate the data bythe T line variable, route_or_line
T_line_alerts <- MonthlyServiceAlertLevelData %>%
  group_by(SubwayLine) %>%
  summarize(total_alerts = sum(distinct_alert_count, na.rm = TRUE))

# plot the sum of alerts by T line

ggplot(T_line_alerts, aes(x = reorder(SubwayLine, -total_alerts), y = total_alerts, fill = SubwayLine)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = scales::label_number(scale = 1e-3, accuracy = 0.1)(total_alerts)), 
            vjust = -0.5, color = "black") + # format labels in millions
  scale_fill_manual(
    values = c(
      "RED" = "red",
      "BLUE" = "blue",
      "GREEN" = "green",
      "ORANGE" = "orange",
      "SILVER" = "darkgray"
    )
  ) + 
  labs(
    title = "Total Service Alerts by MBTA T Line (thousands)",
    subtitle = "Service Alerts from 2021-2024",
    x = "T Line",
    y = "Total Service Alerts"
  ) +
  theme_minimal() +
  theme(legend.position = "none")


```

```{r}
# aggregate the data bythe T line variable, route_or_line
T_line_alerts <- ServiceAlertLevelData %>%
  group_by(SubwayLine) %>%
  summarize(average_severity = mean(severity_code, na.rm = TRUE))

# plot the sum of alerts by T line

ggplot(T_line_alerts, aes(x = reorder(SubwayLine, -average_severity), y = average_severity, fill = SubwayLine)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = scales::label_number(accuracy = 0.1)(average_severity)), 
            vjust = -0.5, color = "black") + # format labels in millions
  scale_fill_manual(
    values = c(
      "RED" = "red",
      "BLUE" = "blue",
      "GREEN" = "green",
      "ORANGE" = "orange",
      "SILVER" = "darkgray"
    )
  ) + 
  labs(
    title = "Average Alert Severity by MBTA T Line",
    subtitle = "Service Alerts from 2021-2024",
    x = "T Line",
    y = "Average Severity of Service Alerts"
  ) +
  theme_minimal() +
  theme(legend.position = "none")


```



```{r} 
# Find the 5 most frequent effect types
top_effects <- ServiceAlertLevelData %>%
  count(effect_name) %>%
  top_n(5, wt = n) %>%
  pull(effect_name)

# Calculate the average severity code for all top 5 effects
top_effects_data <- subset(ServiceAlertLevelData, effect_name %in% top_effects)

# Group and calculate the average severity code
average_severity <- top_effects_data %>%
  group_by(effect_name) %>%
  summarise(average_severity = mean(severity_code, na.rm = TRUE))

ggplot(average_severity, aes(x = reorder(effect_name, -average_severity), y = average_severity, fill = effect_name)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Average Severity Code for Top 5 Effect Types",
    x = "Effect Type",
    y = "Average Severity Code"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_text(aes(label = round(average_severity, 2)), vjust = -0.5, color = "black") # Add labels
```

This next chart shows that Delay alerts are by far the most frequent, 
making them a key area for the MBTA to address. The high frequency of delays 
suggests that focusing on reducing these disruptions could significantly 
improve rider experience. Other alerts like Detour, Elevator Closure,
and Track Change, though less frequent, also impact riders, particularly 
in accessibility and route consistency. To optimize service, MBTA could analyze
these alerts by line or station, differentiate between planned and unplanned 
disruptions, and prioritize maintenance to prevent unexpected outages. For the
coming year, targeted efforts on reducing delays, enhancing communication, and
ensuring accessibility would likely yield the most positive impacts for riders.

```{r}
service_alerts <- ServiceAlertLevelData %>%
  mutate(notif_start = as.Date(notif_start),  # Convert notif_start to Date
         month = floor_date(notif_start, "month"))  # Extract month for aggregation

# Top Alert Frequency
top_alert_types <- ServiceAlertLevelData %>%
  count(effect_name) %>%
  arrange(desc(n)) %>%
  slice(1:5) %>%
  pull(effect_name)

# Only Top Alerts
filtered_alerts <- service_alerts %>%
  filter(effect_name %in% top_alert_types)

# Top Alert Count
monthly_alerts <- filtered_alerts %>%
  group_by(month, effect_name) %>%
  summarise(alert_count = n(), .groups = "drop")

# PAST 3 Years
start_date <- as.Date("2021-01-01")
monthly_alerts <- monthly_alerts %>% filter(month >= start_date)

#VIZ
ggplot(monthly_alerts, aes(x = month, y = alert_count, fill = effect_name)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Monthly Frequency of All Top Alert Types (Last 3 Years)",
       x = "Month",
       y = "Alert Frequency",
       fill = "Alert Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r}
Alert_type <- unique(service_alerts$alert_time_type)

service_alerts %>%
  summarize(total_rows = n(),
            blank_count = sum(alert_time_type == "", na.rm = TRUE),
            na_count = sum(is.na(alert_time_type)))


Alert_counts <- service_alerts %>%
  mutate(alert_time_type = ifelse(alert_time_type == "", NA, alert_time_type)) %>%
  count(alert_time_type, name = "frequency", .drop = FALSE)



ggplot(Alert_counts, aes(x = alert_time_type, y = frequency,  fill= Alert_counts))+
  geom_bar(stat = "Identity", fill = "skyblue", color = "white") +
  labs( title = "Count of Alert Time Type ", x = "Alert Time Type", y = "Count")+
  scale_y_continuous(labels = scales::label_number()) +
  theme_minimal()
```
```{r}

#Looking at gui_mode_name column in service alerts. This column shows the frequency of gui_mode_name, which indicates the type of transport the alert is tied to. Buses have the highest number of alerts. We can explore this further and see what type of alerts are tied to buses as well. 
print(head(service_alerts))

alert_by_mode <- service_alerts %>%
  count(gui_mode_name, name = "frequency")

ggplot(alert_by_mode, aes(x = gui_mode_name, y = frequency, fill = gui_mode_name)) +
  geom_bar(stat = "identity", color = "white")+
  labs( title =" Service Alerts by Transport", x = "Name of Transport", y = "Frequency")+
  scale_y_continuous(labels = scales::label_number())
  theme_minimal()
```


#Created summary data table for subwaylines, looking at the cause code and effect name. Also grouped this data by top 10 subway lines to see what pattern is evident from the type of cause and effect alert name. 
```{r}
 subwayline_summarydata <- service_alerts_subwaylinetagging %>%
  group_by(gui_mode_name,cause_name, cause_code,effect_name,SubwayLine) %>%
  summarize(Total = n(), .groups = "drop") %>%
  arrange(desc(Total)) %>%
  slice_head(n = 10)%>%
tibble()
print(subwayline_summarydata)


ggplot(subwayline_summarydata, aes(x = SubwayLine, y = Total, fill = cause_name ))+
  geom_bar(stat = "identity")+
  labs(title = "Top Alerts by Cause Name and Subway Line",
       x = "Subway Line",
       y = "Total Alerts with Cause Name",
       fill = "Cause Name")+
  theme_minimal()

table(service_alerts_subwaylinetagging$alert_time_type)


grandtotal <- sum(subwayline_summarydata$Total)

sum(monthly_ridership$total_gateentries)

sum(monthly_alerts$alert_count)

count(service_alerts_subwaylinetagging)
```
#This summary shows the top 10 subway alert cause and the type of effect.

### Exploratory Data Analysis for the Merged Dataset with both Ridership & Service Alerts

